---
title: "CPLN505 Assignment II - Assessing Water Infrastructure Need"
author: "Oliver Atwood & Charlie Townsley"
date: "2023-03-01"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    highlight: kate
    code_folding: hide
    code_download: true
---

# Part I: Comparing Groups and Testing for Association and Correlation
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
rm(list=ls())
```

```{r packages data and colors, warning = FALSE, message = FALSE}
#install packages
#only run once
#install.packages("tidyverse")
#install.packages("dplyr")
#install.packages("gt")
#install.packages("ggplot2")
#install.packages("reshape2")
#install.packages("scales")
#install.packages("gmodels")
#install.packages("DT")
#install.packages("corrplot")
#install.packages("Hmisc")
#install.packages("car")

#run every session
library(tidyverse)
library(dplyr)
library(gt)
library(ggplot2)
library(reshape2)
library(scales)
library(gmodels)
library(DT)
library(corrplot)
library(Hmisc)
library(car)

#set colors
palette <- c("#a02300", "#0059c7", "#6496b4", "#E1B93C")
#red and gold are accent
#blue is chart
#dark blue is outline

cwns2012 <- read_csv("https://raw.githubusercontent.com/c-townsley/pbn_a2b_repo/main/Data/cwns2012.csv?token=GHSAT0AAAAAACAEJN5DRTWOHJRNE36CS67GZAZCUFQ")
```


## Memo

**To:** Mayor Kenney  
**From:** Oliver Atwood and Charlie Townsley, O+C Analytics  
**Subject:** The Water Crisis in Philadelphia  

Hey Mayor Kenney, we know you're a busy man, but we've got something important to say! We've got a real problem on our hands, and it's not just about the potholes in the streets or the trash in the alleys. It's about the water.

Using the magic of computers, our team conducted a statistical analysis of national wastewater treatment facility data, and we found that Philadelphia is in desperate need of some serious investment in our water treatment infrastructure. We're talking about \$1.29 billion, Mayor, over $400 million more than the national average! That's a lot of cash.

Now, I know what you're thinking. "Why should I care about this?" Well, let me tell you, Mayor. Without clean water, we've got nothing. No coffee, no beer, no nothing! And that's just the beginning. A lack of investment in water infrastructure can lead to serious health risks, economic losses, and environmental degradation. We’re talking *sewage in the rivers*, Mayor!

We've found that the level of investment needed is correlated with many other values nationally, including population density, low median income, and high projected changes in population density and number of residential households receiving treatment.  Another correlation we unearthed was one of investment need with use of a combined sewer system. Remember that sewage in the rivers I mentioned earlier? Well, if you’re using a combined sewer and it rains, even a little bit, bam! Sewage in the rivers.

Now our analysis has truly been world class, but we need to dig deeper, Mayor. We need to figure out exactly what's going on in our city and come up with a plan to fix it. That's why we're requesting an additional round of research funding to look into this matter further. Specifically, we would like to further explore the relationships between need combined sewer systems, and systems that contribute to highly polluted water bodies. We need to understand the unique challenges we're facing and come up with a solution before it's too late.

So, Mayor Kenney, we’re counting on you to take this seriously. We need to invest in our water infrastructure if we want to have a healthy, thriving city. Let's get to work.

Sincerely,

Oliver Atwood and Charlie Townsley
O+C Analytics

*Update your conclusion to the report to Mayor Kenney summarizing your findings. What additional variables would be helpful to include? What are the major implications of your findings?*

*Highlight conclusion updates*


# Part I.I: Create Metrics to Contextualize Reported Need

We generated the below metrics to measure potential drivers of reported need.
```{r add resilience metrics}
cwns2012 <- cwns2012 %>% 
  mutate(proj_change_res_rec_collctn = PROJ_RES_REC_COLLCTN - PRES_RES_REC_COLLCTN) %>% 
  mutate(proj_prcntchng_res_rec_collctn = (proj_change_res_rec_collctn / PRES_RES_REC_COLLCTN)*100) %>% 
  mutate(proj_change_res_rec_trmt = PROJ_RES_REC_TRMT - PRES_RES_REC_TRMT) %>% 
  mutate(proj_prcntchng_res_rec_trmt = (proj_change_res_rec_trmt / PRES_RES_REC_TRMT)*100) %>% 
  mutate(proj_chng_n_res_rec_collctn = PROJ_N_RES_REC_COLLCTN - PRES_N_RES_REC_COLLCTN) %>% 
  mutate(proj_prcntchng_n_res_rec_collctn = (proj_chng_n_res_rec_collctn / PRES_N_RES_REC_COLLCTN)*100) %>% 
  mutate(proj_chng_n_res_rec_trmt = PROJ_N_RES_REC_TRMT - PRES_N_RES_REC_TRTM) %>% 
  mutate(proj_prcntchng_n_res_rec_trmt = (proj_chng_n_res_rec_trmt / PRES_N_RES_REC_TRTM)*100) %>% 
  mutate(res_pop10_dens = (POP10/(ALAND + AWATER))*1e+6) %>% 
  mutate(res_pop00_dens = (POP00/(ALAND + AWATER))*1e+6) %>%
  mutate(res_pop90_dens = (POP90/(ALAND + AWATER))*1e+6) %>%
  mutate(res_pop80_dens = (POP80/(ALAND + AWATER))*1e+6) %>% 
  mutate(res_pop_dens_chng_80_10 = (res_pop10_dens - res_pop80_dens)) %>% 
  mutate(res_pop_dens_prcnt_chng_80_10 = (res_pop_dens_chng_80_10 / res_pop80_dens)*100) %>%
  mutate(medinc_prcnt_chng_69_09 = ((MEDINC09 - MEDINC69)/MEDINC69)*100)
  
#r clear Nans and Inf values
cwns2012[sapply(cwns2012, is.nan)] <- 0
cwns2012[sapply(cwns2012, is.infinite)] <- 0

#Interactive Summary Table of Resilience Measures by EPA Region
cwns2012 %>% 
  dplyr::select(TOTAL_OFFICIAL_NEED, EPA_REGION, proj_change_res_rec_collctn, proj_prcntchng_res_rec_collctn, 
         proj_change_res_rec_trmt, proj_prcntchng_res_rec_trmt, proj_chng_n_res_rec_collctn, 
         proj_prcntchng_n_res_rec_collctn, proj_chng_n_res_rec_trmt, proj_prcntchng_n_res_rec_trmt,
         res_pop10_dens, res_pop10_dens, res_pop00_dens, res_pop90_dens, res_pop80_dens, 
         res_pop_dens_chng_80_10, medinc_prcnt_chng_69_09) %>% 
  DT::datatable(., options = list(scrollX = TRUE))
```
  

# Part I.II: Summary Statistics of the Data

## Resilience Metrics


**Projected Change in Residential Population Receiving Collection** 
```{r residential population receiving collection}
summary(cwns2012$proj_change_res_rec_collctn)
```


**Projected Percent Change in Residential Population Receiving Collection**
```{r % residential population receiving collection}
summary(cwns2012$proj_prcntchng_res_rec_collctn)
```


**Projected Change in Residential Population Receiving Treatment**
```{r residential population receiving treatment}
summary(cwns2012$proj_change_res_rec_trmt)
```


**Projected Percent Change in Residential Population Receiving Treatment**
```{r % change in residential population receiving treatment}
summary(cwns2012$proj_prcntchng_res_rec_trmt)
```


**Projected Change in Non-Residential Population Receiving Collection**
```{r non-residential population receiving collection}
summary(cwns2012$proj_chng_n_res_rec_collctn)
```


**Projected Percent Change in Non-Residential Population Receiving Collection**
```{r % non-residential population receiving collection}
summary(cwns2012$proj_prcntchng_n_res_rec_collctn)
```


**Projected Change in Non-Residential Population Receiving Treatment**
```{r non-residential population receiving treatment}
summary(cwns2012$proj_chng_n_res_rec_trmt)
```


**Projected Percent Change in Non-Residential Population Receiving Treatment**
```{r % non-residential population receiving treatment}
summary(cwns2012$proj_prcntchng_n_res_rec_trmt)
```


**Residential Population Density 2010** 
*units are persons per kilometer^2^*
```{r Residential population density 2010}
summary(cwns2012$res_pop10_dens)
```


**Change in Population Density 1980 - 2010**
```{r change in res pop density 1980 - 2010}
summary(cwns2012$res_pop_dens_chng_80_10)
```


**Percent Change in Population Density 1980 - 2010**
```{r percent change in res pop density 1980 - 2010}
summary(cwns2012$res_pop_dens_prcnt_chng_80_10)
```


**Percent Change in Median Income 1969 - 2009**
```{r median income 1969 - 2009}
summary(cwns2012$medinc_prcnt_chng_69_09)
```

## Summary Tables

Tables and figures that describe the trends we see in the data, beginning with the top and bottom 10 facilities with the greatest and least need for water infrastructure.
```{r top ten by need}

#Create a new dataframe with the top ten facilities by need
need_top10 <- cwns2012 %>% 
  arrange(desc(TOTAL_OFFICIAL_NEED)) %>% 
  filter(TOTAL_OFFICIAL_NEED != 0) %>% 
  slice(1:10)

#Create a table with the top ten facilities by need and geography
need_top10 %>% 
  dplyr::select(TOTAL_OFFICIAL_NEED, FACILITY_CITY, COUNTYNAME, STATE, EPA_REGION) %>% 
  gt() %>% 
  tab_header(title = "Top Ten Facilities by Need") %>% 
  fmt_currency(columns = c(TOTAL_OFFICIAL_NEED), currency = "USD")

```



```{r bottom ten by need}

#Create a new dataframe with the bottom ten facilities by need
need_bot10 <- cwns2012 %>% 
  arrange(TOTAL_OFFICIAL_NEED) %>% 
  filter(TOTAL_OFFICIAL_NEED != 0) %>% 
  slice(1:10)

#Create a table with the bottom ten facilities by need and geography
need_bot10 %>% 
  dplyr::select(TOTAL_OFFICIAL_NEED, FACILITY_CITY, COUNTYNAME, STATE, EPA_REGION) %>% 
  gt() %>% 
  tab_header(title = "Bottom Ten Facilities by Need") %>% 
  fmt_currency(columns = c(TOTAL_OFFICIAL_NEED), currency = "USD")

```


**Mean Total Official Need**
```{r mean total need}
total_need_mean <- mean(cwns2012$TOTAL_OFFICIAL_NEED)
total_need_mean
```


**Mean Official Need in Philadelphia**
```{r Philadelphia mean official need}
philly_facilities <- cwns2012 %>% 
  filter(STATE == "PA", FACILITY_CITY == "PHILADELPHIA")

need_philly_mean <- mean(philly_facilities$TOTAL_OFFICIAL_NEED)

need_philly_mean
```


**Total Official Need in Philadelphia**
```{r Philadelphia total official need}
need_philly_total <- sum(philly_facilities$TOTAL_OFFICIAL_NEED)

need_philly_total
```


**Difference between Mean Official Need in Philadelphia and National Mean Total Official Need**  
Mean budgeted need of Philadelphia facilities is $407,713,550 higher than mean total official need.
```{r mean philly need less mean national need}
need_philly_mean - total_need_mean
```


**Calculate additional means of density and median income**  
```{r calculate mean density and median income for philadelphia}
density_philly_mean <- mean(philly_facilities$res_pop10_dens)
medincome_philly_mean <- mean(philly_facilities$MEDINC09)
print('Mean Philadelphia Density')
density_philly_mean
print('Mean Philadelphia Median Income')
medincome_philly_mean
```




**Test distribution of log-adjusted Total Official Need**  
*Mean of all facilities in gold*  
*Mean of Philly facilities in red*
```{r Need Histogram}
cwns2012 <- cwns2012 %>% 
  mutate(log_TOTAL_NEED = ifelse(TOTAL_OFFICIAL_NEED > 0, log(TOTAL_OFFICIAL_NEED), TOTAL_OFFICIAL_NEED))

ggplot(cwns2012, aes(x=log_TOTAL_NEED)) +
  geom_histogram(color="#0059c7", size=.1, fill="#6496b4", binwidth=.5, alpha=0.8) +
  geom_vline(aes(xintercept=mean(log_TOTAL_NEED)), color="#E1B93C") +
  geom_vline(aes(xintercept=mean(log(philly_facilities$TOTAL_OFFICIAL_NEED))), color="#a02300") +
  theme_linedraw()

```


**Test distribution of Median Income**  
*Mean of all facilities in gold*  
*Mean of Philly facilities in red*
```{r Income Density Curve}
cwns2012%>%
  ggplot(aes(x=MEANINC09)) +
  ggtitle("Median Income Distribution") +
  xlab("Distribution of Median Income (2009)") + ylab("Median Income") +
  geom_density(fill="#6496b4", size=.35, color="#0059c7", alpha=0.8) +
  geom_vline(aes(xintercept=mean(MEANINC09)), color="#E1B93C") +
  geom_vline(aes(xintercept=mean(philly_facilities$MEANINC09)), color="#a02300") +
  theme_linedraw()

```


## EPA Regions by Resilience Measures

```{r plot of need vs mean density by epa region}
#Summarize total need and group by EPA region
mean_density_need_epa <- cwns2012 %>%
  group_by(EPA_REGION) %>% 
  summarise(count = n(), mean_popdens = mean(res_pop10_dens), mean_need = mean(TOTAL_OFFICIAL_NEED))

ggplot(mean_density_need_epa, aes(x = mean_popdens, y = mean_need)) +
  geom_smooth(method=lm, color="#0059c7") +
  geom_point(size=3, color="black", fill="#6496b4", shape=21, alpha=.5) +
  ggtitle("Population Density vs. Total Need, by EPA Region") +
  geom_text(aes(label = rownames(mean_density_need_epa), hjust = 1.75, vjust = 1)) +
  xlab("Mean Population Density (2010)") + ylab("Mean Official Need") +
  theme_linedraw()
  
```

Density and need seem to be correlated. So how does the mean need of Philadelphia facilities compare with the mean values for each EPA region?

Here is the same plot with Philadelphia shown as a red dot.
```{r plot of need vs mean density by epa region plus philadelphia point, message=FALSE}
mean_density_need_epa <- cwns2012 %>%
  group_by(EPA_REGION) %>% 
  summarise(count = n(), mean_popdens = mean(res_pop10_dens), mean_need = mean(TOTAL_OFFICIAL_NEED))

ggplot(mean_density_need_epa, aes(x = mean_popdens, y = mean_need)) +
  geom_smooth(method=lm, color="#0059c7") + 
  geom_point(size=3, color="black", fill="#6496b4", shape=21, alpha=.5) +
  ggtitle("Population Density vs. Total Need, by EPA Region") +
  geom_text(aes(label = rownames(mean_density_need_epa), hjust = 1.75, vjust = 1)) +
  xlab("Mean Population Density (2010)") + ylab("Mean Official Need") +
  geom_point(aes(x = density_philly_mean , y = need_philly_mean),size=3, shape=21,color="black", fill="#a02300") +
  theme_linedraw()

```


How does Philly's density and need compare to other cities in the same EPA region?
```{r plot of need vs mean density in epa region 3}
#Summarize total need in EPA region 3
mean_density_need_R3 <- cwns2012 %>%
  filter(EPA_REGION == '3') %>% 
  group_by(FACILITY_CITY) %>% 
  summarise(count = n(), mean_popdens = mean(res_pop10_dens), mean_need = mean(TOTAL_OFFICIAL_NEED))

ggplot(mean_density_need_R3, aes(x = mean_popdens, y = mean_need)) +
  geom_point(size=3, color="black", fill="#6496b4", shape=21, alpha=.5) +
  ggtitle("Population Density vs. Total Need by City in EPA Region 3") +
  xlab("Mean Population Density (2010)") + ylab("Mean Official Need") +
  geom_point(aes(x = density_philly_mean , y = need_philly_mean),size=3, shape=21,color="black", fill="#a02300") +
  theme_linedraw()

```


What about income and need in the same EPA region?
```{r plot of need vs median income in epa region 3}
#Summarize total need in EPA region 3
mean_income_need_R3 <- cwns2012 %>%
  filter(EPA_REGION == '3') %>% 
  group_by(FACILITY_CITY) %>% 
  summarise(count = n(), mean_medinc = mean(MEDINC09), mean_need = mean(TOTAL_OFFICIAL_NEED))

ggplot(mean_income_need_R3, aes(x = mean_medinc, y = mean_need)) +
  geom_point(size=3, color="black", fill="#6496b4", shape=21, alpha=.5) +
  ggtitle("Median Income vs. Total Need by City in EPA Region 3") +
  xlab("Median Income (2009)") + ylab("Mean Official Need") +
  geom_point(aes(x = medincome_philly_mean , y = need_philly_mean),size=3, shape=21,color="black", fill="#a02300") +
  theme_linedraw()

```




# Part I.III: Tests of Association

## Comparing Budgeted Need for Facilities with CSS vs MS4s


Summary of overall budgeted need for facilities with CSS (Combined Sewer Systems):
```{r css need}
cwns2012_css <- filter(cwns2012, CSS == 1)
summary(cwns2012_css$TOTAL_OFFICIAL_NEED)
```


Summary of overall budgeted need for facilities with MS4s (Municipal Separate Storm Sewer Systems):
```{r ms4 need}
cwns2012_ms4 <- filter(cwns2012, MS4 == 1)
summary(cwns2012_ms4$TOTAL_OFFICIAL_NEED)
```




We can see that mean budgeted need for facilities with CSS is higher than for facilities with MS4s, but is this difference statistically meaningful?

Conducting a Welch two-sample t-test reveals that these means are significantly different from each other.

It gives a t-statistic that is greater than the critical value for significance of 1.96 and allows us reject the null hypothesis that the mean budgeted need is the same for CSS and MS4s with greater than 95% confidence.
```{r two-sample t-test css vs ms4}
t.test(TOTAL_OFFICIAL_NEED ~ CSS, data = cwns2012, paired = FALSE)
```




## Comparing Need by Association with a Total Maximum Daily Load (TMDL) of one or more pollutants


Budgeted need for facilities that contribute to a TMDL receiving water body:
```{r yes tmdl}
cwns_yestmdl <- filter(cwns2012, TMDL_INDICATOR == "Y")
summary(cwns_yestmdl$TOTAL_OFFICIAL_NEED)
```


Budgeted need for facilities that do not contribute to a TMDL receiving water body:
```{r no tmdl}
cwns_notmdl <- filter(cwns2012, TMDL_INDICATOR == "N")
summary(cwns_notmdl$TOTAL_OFFICIAL_NEED)
```


Mean budgeted need for facilities that do contribute to a TMDL is higher than for facilities that do not.

Running a Welch two sample t-test shows that mean budgeted need is statistically different between facilities that do and do not contribute to a TMDL.

It gives a t-statistic that surpasses the critical value for significance of 1.96 and allows us reject the null hypothesis that the means are the same with greater than 95% confidence.
```{r two-sample t-test tmdl}
t.test(TOTAL_OFFICIAL_NEED ~ TMDL_INDICATOR, data = cwns2012, paired = FALSE)
```




## Comparing Significance of Difference between Facilities with and without TMDL by EPA Region

We found a statistically significant difference in mean budgeted need between facilities that do contribute to a TMDL and those that do not in just four out of ten EPA Regions. These were regions 2, 3, 5, and 6.

Because we could not reject the null hypothesis (that means are different) in 60% of EPA regions, we do not see TMDL by EPA region as a useful metric for further exploration.


**Region 2**
```{r two-sample t-test tmdl for epa region 2}
t.test(TOTAL_OFFICIAL_NEED ~ TMDL_INDICATOR, data = filter(cwns2012, EPA_REGION == 2), paired = FALSE)
```


**Region 3**
```{r two-sample t-test tmdl for epa region 3}
t.test(TOTAL_OFFICIAL_NEED ~ TMDL_INDICATOR, data = filter(cwns2012, EPA_REGION == 3), paired = FALSE)
```


**Region 5**
```{r two-sample t-test tmdl for epa region 5}
t.test(TOTAL_OFFICIAL_NEED ~ TMDL_INDICATOR, data = filter(cwns2012, EPA_REGION == 5), paired = FALSE)
```


**Region 6**
```{r two-sample t-test tmdl for epa region 6}
t.test(TOTAL_OFFICIAL_NEED ~ TMDL_INDICATOR, data = filter(cwns2012, EPA_REGION == 6), paired = FALSE)
```

We found no statistically significant difference (failed to reject the null hypothesis) in mean budgeted need between facilities that do contribute to a TMDL and those that do not in regions 1, 4, 9, and 10.


**Region 1**
```{r two-sample t-test tmdl for epa region 1}
t.test(TOTAL_OFFICIAL_NEED ~ TMDL_INDICATOR, data = filter(cwns2012, EPA_REGION == 1), paired = FALSE)
```


**Region 4**
```{r two-sample t-test tmdl for epa region 4}
t.test(TOTAL_OFFICIAL_NEED ~ TMDL_INDICATOR, data = filter(cwns2012, EPA_REGION == 4), paired = FALSE)
```


**Region 9**
```{r two-sample t-test tmdl for epa region 9}
t.test(TOTAL_OFFICIAL_NEED ~ TMDL_INDICATOR, data = filter(cwns2012, EPA_REGION == 9), paired = FALSE)
```


**Region 10**
```{r two-sample t-test tmdl for epa region 10}
t.test(TOTAL_OFFICIAL_NEED ~ TMDL_INDICATOR, data = filter(cwns2012, EPA_REGION == 10), paired = FALSE)
```


**Regions 7 and 8**  
Regions 7 and 8 both failed the t-test outright because there was not enough data.
```{r two-sample t-test tmdl for epa region 7 and 8, error=TRUE}
# t.test(TOTAL_OFFICIAL_NEED ~ TMDL_INDICATOR, data = filter(cwns2012, EPA_REGION == 7), paired = FALSE)

# t.test(TOTAL_OFFICIAL_NEED ~ TMDL_INDICATOR, data = filter(cwns2012, EPA_REGION == 8), paired = FALSE)
```




## Chi-square Test of Regional Differences in Total Official Need

To test for association between total budgeted need and EPA region, we first need to divide total need into a limited number of categories that a cross table can process.
```{r breaking total offical need into 6 classes}
#First create a new column that breaks total official need into six classes
cwns_int <- cwns2012 %>%
  mutate(need_int = ntile(cwns2012$TOTAL_OFFICIAL_NEED, 5)) %>% 
  dplyr::select(need_int, TOTAL_OFFICIAL_NEED)
```

Chi-Square test of association between need and EPA region shows that we can confidently reject the null hypothesis that need and epa region are independent variables due to the incredibly small p-value.
```{r cross table of total official need by epa region}
CrossTable(cwns2012$EPA_REGION, cwns_int$need_int,  fisher = FALSE, chisq = TRUE, 
             expected = TRUE, sresid = FALSE, format="SPSS")
```




## Testing 5 Variables for Association

Below are five variables from those we originally explored, divided into five equal intervals and tested for their association with total reported need.


**2010 residential population density is associated with total official need.**
```{r cross table of association between total official need and variable 1}
#First create a new column that breaks res_pop10_dens into distinct classes
#1 is low and 5 is high
cwns_int <- cwns_int %>%
  mutate(res_pop10_dens = ntile(cwns2012$res_pop10_dens, 5))

#cross tab of res_pop10_dens and need
CrossTable(cwns_int$need_int, cwns_int$res_pop10_dens,  fisher = FALSE, chisq = TRUE, 
             expected = TRUE, sresid = FALSE, format="SPSS")
```


**Residential population density change from 1980 - 2010 is associated with total official need.**
```{r cross table of association between total official need and res pop dens chng 1980 - 2010}
#First create a new column that breaks res_pop10_dens into six classes
cwns_int <- cwns_int %>%
  mutate(res_pop_dens_chng_80_10 = ntile(cwns2012$res_pop_dens_chng_80_10, 5))

#cross tab of res_pop10_dens and need
CrossTable(cwns_int$need_int, cwns_int$res_pop_dens_chng_80_10,  fisher = FALSE, chisq = TRUE, 
             expected = TRUE, sresid = FALSE, format="SPSS")
```


**2010 population size is associated with total official need.**
```{r cross table of association between need and pop10}
#First create a new column that breaks res_pop10_dens into six classes
cwns_int <- cwns_int %>%
  mutate(POP10 = ntile(cwns2012$POP10, 5))

#cross tab of res_pop10_dens and need
CrossTable(cwns_int$need_int, cwns_int$POP10,  fisher = FALSE, chisq = TRUE, 
             expected = TRUE, sresid = FALSE, format="SPSS")
```


**Projected change in residential receiving treatment is associated with total official need.**
```{r cross table of association between need and proj_change_res_rec_trmt}
#First create a new column that breaks res_pop10_dens into six classes
cwns_int <- cwns_int %>%
  mutate(proj_change_res_rec_trmt = ntile(cwns2012$proj_change_res_rec_trmt, 5))

#cross tab of res_pop10_dens and need
CrossTable(cwns_int$need_int, cwns_int$proj_change_res_rec_trmt,  fisher = FALSE, chisq = TRUE, 
             expected = TRUE, sresid = FALSE, format="SPSS")
```


**Projected change in residential receiving collection is associated with total official need.**
```{r cross table of association between need and proj_change_res_rec_collctn}
#First create a new column that breaks the variable into six classes
cwns_int <- cwns_int %>%
  mutate(proj_change_res_rec_collctn = ntile(cwns2012$proj_change_res_rec_collctn, 5))

#cross tab of var5 and need
CrossTable(cwns_int$need_int, cwns_int$proj_change_res_rec_collctn,  fisher = FALSE, chisq = TRUE, 
             expected = TRUE, sresid = FALSE, format="SPSS")
```




# Part I.IV: Correlation

Correlation tests were run for all the resilience measures, as well as for 2010 population, 2010 population density, 2009 median income, and present residential receiving treatment.
Values with insignificant correlation with total official need were commented out of the code.

### Correlation of Variables of Interest with Total Official Need

A number of variables exhibited no significant correlation with total official need at all. Expand the code block to see them.
```{r insignificant Correlation Coefficient Tests}
#cor(cwns2012$medinc_prcnt_chng_69_09, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
#cor(cwns2012$res_pop_dens_prcnt_chng_80_10, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
#cor(cwns2012$proj_prcntchng_res_rec_trmt, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
#cor(cwns2012$proj_prcntchng_res_rec_collctn, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
#cor(cwns2012$proj_prcntchng_n_res_rec_trmt, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
#cor(cwns2012$proj_prcntchng_n_res_rec_collctn, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
#cor(cwns2012$MEDINC09, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
```


Similarly, median income seems to have an extremely weak correlation with need based on its correlation coefficient.
```{r correlation medinc09}
cor(cwns2012$MEDINC09, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
```
Projected change in non-residential population receiving treatment has a slight correlation with need.
```{r correlation proj chng res rec trtmt}
cor(cwns2012$proj_chng_n_res_rec_trmt, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
```
Projected change in non-residential population receiving collection has nearly the same slight correlation (0.1% higher).
```{r correlation proj chng n res rec collctn}
cor(cwns2012$proj_chng_n_res_rec_collctn, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
```

2010 population is in the same category of slight correlation.
```{r correlation pop10}
cor(cwns2012$POP10, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
```

Projected change in residential population receiving collection has a more notable correlation with need, although it is still small.
```{r correlation proj chng res rec collctn}
cor(cwns2012$proj_change_res_rec_collctn, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
```
Projected change in residential population receiving treatment has a very similar notable but small correlation.
```{r correlation proj chng res rec trmt}
cor(cwns2012$proj_change_res_rec_trmt, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
```
Change in residential population density from 1980 to 2010 has a moderate correlation with need that is notably higher than the above variables.
```{r correlation res pop dens chng}
cor(cwns2012$res_pop_dens_chng_80_10, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
```
Although residential population density in 2010 still exhibits a moderate correlation with need overall, it's correlation coefficient is significantly higher than the rest of the variables we tested.
```{r correlation res pop10 dens}
cor(cwns2012$res_pop10_dens, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
```
```{r correlation res pop10 dens}
cor(cwns2012$PRES_RES_REC_TRMT, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
```

```{r correlation res pop10 dens}
cor(cwns2012$PRES_RES_REC_COLLCTN, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
```

```{r correlation res pop10 dens}
cor(cwns2012$PCTWHITE10, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
```

```{r correlation res pop10 dens}
cor(cwns2012$CSS, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
```


## Scatter Plots

For the 8 values where correlation was found, log-adjusted scatter plots were generated.
```{r Scatter Plot res pop dens Change 1980 - 2010 by Need}

ggplot(cwns2012, aes(x=log(res_pop_dens_chng_80_10), y=log(TOTAL_OFFICIAL_NEED))) + 
  ggtitle("Residential Population Density Change by Official Need") +
  xlab("Log-Adjusted Residential Population Density Change (1980 - 2010)") + ylab("Log-Adjusted Official Need") +
  geom_point(size=1, color="black", fill="#6496b4", shape=21, alpha=.5) +
  geom_smooth(method=lm, color="#0059c7") +
  theme_linedraw()

```
```{r Scatter Plot proj chng in res pop rec trtmnt by Need}

ggplot(cwns2012, aes(x=log(proj_change_res_rec_trmt), y=log(TOTAL_OFFICIAL_NEED))) + 
  ggtitle("Projected Change in Res. Population Receiving Treatment by Official Need") +
  xlab("Log-Adjusted Projected Change in Residential Population Receiving Treatment") + ylab("Log-Adjusted Official Need") +
  geom_point(size=1, color="black", fill="#6496b4", shape=21, alpha=.5)+
  geom_smooth(method=lm, color="#0059c7") +
  theme_linedraw()

```
```{r Scatter Plot of proj chng in res pop rec collctn by Need}

ggplot(cwns2012, aes(x=log(proj_change_res_rec_collctn), y=log(TOTAL_OFFICIAL_NEED))) + 
  ggtitle("Projected Change in Res. Population Receiving Collection by Official Need") +
  xlab("Log-Adjusted Projected Change in Residential Population Receiving Collection") + ylab("Log-Adjusted Official Need") +
  geom_point(size=1, color="black", fill="#6496b4", shape=21, alpha=.5)+
  geom_smooth(method=lm, color="#0059c7") +
  theme_linedraw()

```
```{r Scatter Plot of proj chng in n-res pop rec trtmnt by Need}

ggplot(cwns2012, aes(x=log(proj_chng_n_res_rec_trmt), y=log(TOTAL_OFFICIAL_NEED))) + 
  ggtitle("Projected Change in Non-Res. Population Receiving Treatment by Official Need") +
  xlab("Log-Adjusted Projected Change in Non-Residential Population Receiving Treatment") + ylab("Log-Adjusted Official Need") +
  geom_point(size=1, color="black", fill="#6496b4", shape=21, alpha=.5)+
  geom_smooth(method=lm, color="#0059c7") +
  theme_linedraw()

```
```{r Scatter Plot of proj chng in N-res pop rec collctn by Need}

ggplot(cwns2012, aes(x=log(proj_chng_n_res_rec_collctn), y=log(TOTAL_OFFICIAL_NEED))) + 
  ggtitle("Projected Change in Non-Res. Population Receiving Collection by Official Need") +
  xlab("Log-Adjusted Projected Change in Non-Residential Population Receiving Collection") + ylab("Log-Adjusted Official Need") +
  geom_point(size=1, color="black", fill="#6496b4", shape=21, alpha=.5)+
  geom_smooth(method=lm, color="#0059c7") +
  theme_linedraw()

```
```{r Scatter Plot of res pop dens 2010 by Need}

ggplot(cwns2012, aes(x=log(res_pop10_dens), y=log(TOTAL_OFFICIAL_NEED))) + 
  ggtitle("Residential Population Density (2010) by Official Need") +
  xlab("Log-Adjusted Residential Population Density in 2010") + ylab("Log-Adjusted Official Need") +
  geom_point(size=1, color="black", fill="#6496b4", shape=21, alpha=.5)+
  geom_smooth(method=lm, color="#0059c7") +
  theme_linedraw()

```
```{r Scatter Plot of Residential pop 2010 by Need}

ggplot(cwns2012, aes(x=log(POP10), y=log(TOTAL_OFFICIAL_NEED))) + 
  ggtitle("Residential Population (2010) by Official Need") +
  xlab("Log-Adjusted Residential Population (2010)") + ylab("Log-Adjusted Official Need") +
  geom_point(size=1, color="black", fill="#6496b4", shape=21, alpha=.5)+
  geom_smooth(method=lm, color="#0059c7") +
  theme_linedraw()

```
```{r Scatter Plot of med Income in 2009 by Need}

ggplot(cwns2012, aes(x=log(POP10), y=log(TOTAL_OFFICIAL_NEED))) + 
  ggtitle("Median Income (2009) by Official Need") +
  xlab("Log-Adjusted Median Income (2009)") + ylab("Log-Adjusted Official Need") +
  geom_point(size=1, color="black", fill="#6496b4", shape=21, alpha=.5)+
  geom_smooth(method=lm, color="#0059c7") +
  theme_linedraw()

```

# Part IIb

## Data Preparation and Preliminary Tests

### Create Variables to Represent Need

> From the homework, delete when done:

> Suspecting that Philadelphia’s high infrastructure need can be operationalized in different ways, you create the following variables to represent ‘need’:
• Total Official Need ($, untransformed)
• Log of Total Official Need (log $) 
• Residential Burden ($/person) 
• Log of Residential Burden (log $/person)

> Give an explanation for why you choose to operationalize ‘need’ in the above ways (what each can reveal, and the interpretations of coefficients from regression change). Boxplots and density/histograms of these variables can help illustrate very skewed distributions.

> Hint: Be careful of dividing by zeros in communities that have either reported zero residential treatment or zero residential collection values. You can assume that most residents that receive treatment also receive collection and vice versa, but that non-standard reported has resulted in some inconsistency in the data.

To model infrastructure need, we need variables that represent it. From our tests of correlation and association, we select the following variables (based on XX) to include in our models:

* res_pop10_dens  
* res_pop_dens_chng_80_10 
* proj_change_res_rec_trmt
* proj_change_res_rec_collctn
* POP10
* MEDINC09
* PRES_RES_REC_TRMT
* PRES_RES_REC_COLLCTN
* PCTWHITE10
* CSS

To better understand Philadelphia's especially high infrastructure need, we create additional variables:

* ResBurden  
* LogResBurden  
* LogOfficialNeed  
* LogPOP10
* LogMEDINC09

Calculating Additional Variables:

```{r Calculate Residential Burden}
cwns2012 <- cwns2012 %>% 
  mutate(ResBurden = TOTAL_OFFICIAL_NEED/max(PROJ_RES_REC_COLLCTN, PROJ_RES_REC_TRMT)) %>% 
  mutate(LogResBurden = ifelse(ResBurden > 0, log(ResBurden), 0)) %>% 
  mutate(LogPOP10 = ifelse(POP10 > 0, log(POP10), 0)) %>% 
  mutate(LogMEDINC09 = ifelse(MEDINC09 > 0, log(MEDINC09), 0))
```


### Creating Binary "Dummy" Variables

> From the homework, delete when done:
> Dummy variables for ‘groups’ such as ‘growing cities’ or ‘cities with non-residential populations’

> You could also try to incorporate dummy variables for any other categorizations you think would explain facilities’ systematically higher or lower reported needs. Remember: dummy variables are essentially intercept adjustments in a regression that are based on the entire group having a higher/lower mean than non-group facilities. From what you found from your previous scope (i.e. any significant associations or correlations you found in Part I. of the assignment), you may also wish to include some of the change and percent change metrics you calculated to explain ‘need’. Explain theoretical reasons why you want to include certain variables in your regressions.

Suspecting that some factors influencing waste water infrastructure need in Philadelphia may be categorical rather than continuous, we create the following binary dummy variables to include in our models:

* GrowingCity
* high income?

```{r Dummy variable creation}

cwns2012 <- cwns2012 %>% 
  mutate(GrowingCity = ifelse(POP10-POP80 > 0, 1, 0)) %>% 
  mutate(highincome = ifelse(MEDINC09 >= ))

```

### Testing for multi-coliniearity

Final variable list to test:
* res_pop10_dens  
* res_pop_dens_chng_80_10 
* proj_change_res_rec_trmt
* proj_change_res_rec_collctn
* POP10
* MEDINC09
* PRES_RES_REC_TRMT
* PRES_RES_REC_COLLCTN
* PCTWHITE10
* ResBurden  
* LogResBurden  
* LogOfficialNeed  
* LogPOP10
* LogMEDINC09
* CSS
* GrowingCity


Running pairwise correlations between variables (visualized with the correlogram below) revealed that some of the variables we are interested in testing are highly correlated.  

The following variables exhibited high degrees of significance with each other:  
* log of official need & log of residential burden  **(add cor. coefficient)** so we cannot use residential burden in our models.
* projected change in residences receiving treatment & projected change in residences receiving collection **(add cor. coefficient)**. We chose to move forward with projected change in residences receiving treatment because the corr. score is **XX** points higher for treatment.

* residential population density in 2010 & residential population density change from 1980 to 2010 **(add cor. coefficient)**. We chose to use residential population in 2010 in 2010 moving forward because it has a higher correlation with total official need.


```{r correlation matrix}
#make matrix of variables we're testing
mod_vars <- cwns2012 %>% 
  dplyr::select(TOTAL_OFFICIAL_NEED, log_TOTAL_NEED, res_pop10_dens, res_pop_dens_chng_80_10,
                proj_change_res_rec_trmt, proj_change_res_rec_collctn, POP10, MEDINC09,
                ResBurden, LogResBurden, PRES_RES_REC_TRMT, PRES_RES_REC_COLLCTN, LogPOP10, LogMEDINC09, PCTWHITE10, CSS, GrowingCity)

#compute correlation matrix
cormatrix <- cor(mod_vars) %>% 
  round(., 2)

#plot a correlogram
corrplot(cormatrix, method = "circle", type = "lower", order = "hclust", 
         tl.col = "black", tl.srt = 45)
```
```{r correlations}
print('res_pop10_dens')
cor(cwns2012$res_pop10_dens, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
print('res_pop_dens_chng_80_10')
cor(cwns2012$res_pop_dens_chng_80_10, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
print('proj_change_res_rec_trmt')
cor(cwns2012$proj_change_res_rec_trmt, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
print('proj_change_res_rec_collctn')
cor(cwns2012$proj_change_res_rec_collctn, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
print('POP10')
cor(cwns2012$POP10, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
print('LogPOP10')
cor(cwns2012$LogPOP10, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
print('PRES_RES_REC_TRMT')
cor(cwns2012$PRES_RES_REC_TRMT, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
print('PCTWHITE10')
cor(cwns2012$PCTWHITE10, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
print('ResBurden')
cor(cwns2012$ResBurden, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
print('LogResBurden')
cor(cwns2012$LogResBurden, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
print('LogOfficialNeed')
cor(cwns2012$TOTAL_OFFICIAL_NEED, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
print('LogMEDINC09')
cor(cwns2012$LogMEDINC09, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
print('CSS')
cor(cwns2012$CSS, cwns2012$TOTAL_OFFICIAL_NEED, use="complete.obs", method="pearson")
```


Based on correlation with Total Official Need, and elimination of one of a variable pair when colinearity is present, favoring the one with higher correlation with need, we narrowed down this list of variables to 5 indie vars.

Final variable list to test:
* res_pop10_dens (cor with need = 0.42)
* PRES_RES_REC_TRMT (cor with need = 0.53)
* LogResBurden (cor with need = 0.39)
* CSS (cor with need = 0.20)
* PCTWHITE10 (cor with need = -0.15)

## Bivariate Regressions

> From the homework, delete when done:

> Run bivariate regressions of explanatory variables against your ‘need variables.’ Make scatterplots
showing the linear relationships between them (convention is to put the independent variable on
the x-axis and the dependent variable, ‘need’ in this case, on the y-axis). Interpret the coefficient
estimates of significant variables. How well do these simple bivariate models perform in
predicting the various measures of ‘need’?

Testing relationships between each of the variables considered above (minus the variables that exhibit strong colinearity).

Testing relationships between three of the variables considered above which have the highest correlations with need.

Need vs Residential Population Density 2010
```{r bivariate regression 1}
bimod1 <- lm(TOTAL_OFFICIAL_NEED ~ res_pop10_dens, data = cwns2012)
summary(bimod1)

plot(fitted(bimod1), bimod1$residuals,
     main = "Fitted Model Against Residuals: Population Density",
     xlab = "Fitted Model 1", ylab = "Residuals",
     col="#6496b4", pch = 16)
abline(0,0, col = "#0059c7")

```
Want the points to be symmetric above and below the line (homoschedasticity assumption > we want equal variances). This doesn't look ideal because more of the residuals appear to be above the line.
Also want to avoid linearity.
Google ideal residual plots.
Some of the assumptions of linear regression are not met here with regards to homoschedasticity and linearity.


```{r bivariate regression 2}
bimod2 <- lm(TOTAL_OFFICIAL_NEED ~ PRES_RES_REC_TRMT, data = cwns2012)
summary(bimod2)

plot(fitted(bimod2), bimod2$residuals,
     main = "Fitted Model Against Residuals: Residential Receiving Treatment",
     xlab = "Fitted Model 2", ylab = "Residuals",
     col="#6496b4", pch = 16)
abline(0,0, col = "#0059c7")
```

```{r bivariate regression 3}
bimod3 <- lm(TOTAL_OFFICIAL_NEED ~ LogResBurden, data = cwns2012)
summary(bimod3)

plot(fitted(bimod3), bimod3$residuals,
     main = "Fitted Model Against Residuals: Log Residential Burden",
     xlab = "Fitted Model 3", ylab = "Residuals",
     col="#6496b4", pch = 16)
abline(0,0, col = "#0059c7")
```
This is happening because looking at need vs. log of need. Don't include this model.


```{r bivariate regression 4}
bimod4 <- lm(TOTAL_OFFICIAL_NEED ~ CSS, data = cwns2012)
summary(bimod4)

plot(fitted(bimod4), bimod4$residuals,
     main = "Fitted Model Against Residuals: Combined Sewer System",
     xlab = "Fitted Model 4", ylab = "Residuals",
     col="#6496b4", pch = 16)
abline(0,0, col = "#0059c7")
```


```{r bivariate regression 5}
bimod5 <- lm(TOTAL_OFFICIAL_NEED ~ PCTWHITE10, data = cwns2012)
summary(bimod5)

plot(fitted(bimod5), bimod5$residuals,
     main = "Fitted Model Against Residuals: Percent White Population",
     xlab = "Fitted Model 5", ylab = "Residuals",
     col="#6496b4", pch = 16)
abline(0,0, col = "#0059c7")
```
Residual Plots for each of these 5 variables. 



## Multivariate Regression

> From the homework, delete when done:

> Build multivariate regressions for each of the measures of ‘need’. Explain the use of, and report
the results of automated variable selection techniques, such as Forward and Backward Selection.

> In your final models for each of the measures of ‘need’, report which variables you chose to
include in your final “leanest and meanest” models, interpret the significant coefficients of the
final models. Make sure that you interpret interesting results in full sentences, as well as
presenting regression results in a table summarizing your work. Compare models against each
other, using R squared values and F-tests.

> Remember to take care in interpreting log-transformed variables.
Try interactions between variables. Do any have significant effects?
Report the VIF values to convince suspicious readers that you have considered the problem of
multicollinearity in your analysis.

Based on the results of the correlogram and bivariate regresssions, we create the following variable sets to test with multivariate regressions. We create multiple variable sets to independently test variables that exhibit multi-colinearity but still appear to be significant.

**Variable Set 1**  
* TOTAL_OFFICIAL_NEED  
* res_pop10_dens
* proj_change_res_rec_trmt (the treatment variable has a **slightly** higher correlation with need than collection)
* POP10
* MEDINC09
* LogResBurden
* CSS  
* GrowingCity

### Using Backward Selection to Select Variables

**Model 1**
For our first model, we begin with a selection of all of relevant variables that do not exhibit multi-collinearity according to our correlation tests.

```{r multivariate regresssion model 1}
#Interpreting regression results - key statistics:
#R-squared
  # Tells what % of the variability in the dependent variable is explained by the model
  # in other words - r-squared measures the strength of the relationship btwn the model
  # and dependent variable on a 0-100% scale
  # so, higher r-squared values represent smaller differences btwn the observed data
  # and the fitted values.
  # r-squared < 50% is common when studying human behavior
  # IMPORTANT NOTE: r-squared increases every time you add an independent variable to the model.
#Adjusted R-squared
  # Adjusts for the number of terms in the model!
  # Its value increases only when the new term improves the model fit more than expected by chance alone.
  # The adjusted R-squared value actually decreases when the term doesn’t improve the model fit enough.
  # Useful for comparing goodness-of-fit for regression models with differing numbers of independent variables.
#Multiple R-squared
#Significance codes
#F-statistic
  # a large f-stat means the model has a strong fit and we can reject the
  # null hypothesis that a null model (just the intercept) describes the data better.
  # If the p-value associated with the F-stat is ≥ 0.05 Then there is no relationship between ANY of the
  # independent variables and Y.
  # If the p-value associated with the F-statistic < 0.05 then AT LEAST 1 independent variable is related to Y.
#AIC - oliver to fill out
  # Particularly helpful for backwards selection

mod1.0 <- lm (TOTAL_OFFICIAL_NEED ~ POP10 + res_pop10_dens + MEDINC09 + proj_change_res_rec_trmt + LogResBurden + CSS + GrowingCity, data = cwns2012)

summary(mod1.0)
```

**Interpret Results of mod1**
R-squared and adjusted R-squared aren't awful but they could be higher. Median income doesn't seem to be significant. F-statistic p-value is low. Every other variable appears to be significant. Model 1 could perform better, but seems like a good start.

**Test mod1 for multi-colinearity**
The VIF for Model 1 reveals that multiple variables exhibit multiple co-linearity. 2010 Population, 2010 residential population density, the interaction between the two, and the interaction between 2010 population and 2009 median income all have VIF values above 5 and square root VIF values above 2.

```{r multi-colinearity test mod1}
#variance inflation factor (VIF)
  # Index of how a coefficient variance increases due to colinearity.
  # VIF calculates how correlated the variable is with all the other variables in the model.
  # Mechanics: Regress X1 on X2, calculate r2 of this relationship
  # Calculate the VIF for X1 as follows:  1/(1 - r2)
  # some people say if VIF>5-10 then you have a problem, 
  # others are more conservative and say if sqrt(vif)>2 then you have a problem

vif(mod1.0) # variance inflation factors 
sqrt(vif(mod1.0))
```

All of our VIFs are less than 2, so we can proceed with confidence that colinearity among these variables is not hindering the model.

**Model 2**
Now we create a new model based on these results. Growing city dummy variable was removed due to low significance.

```{r mod 2}
mod1.1 <- lm (TOTAL_OFFICIAL_NEED ~ POP10 + res_pop10_dens + MEDINC09 + proj_change_res_rec_trmt + LogResBurden:CSS, data = cwns2012)

summary(mod1.1)
```

**Test mod1 for multi-colinearity**

```{r multi-colinearity test2 mod1}
vif(mod1.1) # variance inflation factors 
sqrt(vif(mod1.1))
```

## Final Model

Why we choose this model.

Describe coefficients (for each 1 unit increase in these variables, we expect the following change in total need).

Provide diagnosis of residuals vs. fitted

Interpretation: what factors relate to total official need? How does they relate to total official need in Philly?

What does this tell us about wastewater facility need nationally and in Philadelphia?


